\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-24 response.tex

\documentclass[12pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[margin=20mm]{geometry}
\usepackage{amsmath,color,amsthm,enumerate,amsfonts,mathtools}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{todonotes}

\newcommand{\done}{\textcolor{blue}{Done.}}
\newcommand{\changed}{\textcolor{blue}{Revised accordingly.}}
\newcommand{\tba}{\textcolor{red}{To be addresssed.}}

@@

\setlength{\parindent}{0ex}
\setlength{\parskip}{1ex}

\begin{document}

\subsection*{Graph Product Structure for Non-Minor-Closed Classes\\
Vida Dujmovi\'c, Pat Morin, David R. Wood\\
Journal of Combinatorial Theory Series B (JCTB9758)\\
Response to Referee Reports}

Both referees strongly recommend acceptance. We have revised the paper according to the referee reports as detailed below.

@@

\subsection*{Referee \#1}

%This paper proves that if a class of graphs F admits a product structure
%theorem, then the class of all graphs which have a 1-subdivision in F admits
%a product structure theorem as well. The main theorem, Theorem 3, while
%equivalent to the above statement using standard techniques, is stated in
%a different way which is considerably more useful for applications and for
%obtaining good bounds. As applications of Theorem 3, the paper shows that
%the following classes admit a product structure theorem:
%
%\begin{itemize}
%\item graphs drawn on a fixed surface with each edge in $\leq k$ crossings,
%\item map graphs of degree $\leq k$ graphs embedded in a fixed surface,
%\item string graphs with each curve in $\leq k$ intersections,
%\item $k$-nearest neighbour graphs in the plane, and
%\item powers of classes of graphs of degree $\leq k$ which admit a product structure theorem.
%\end{itemize}
%
%All but the last bullet point are shown to follow from the first.
%The main results in the paper are correct, strong, and of high interest,
%and the paper should definitely be accepted to JCTB after some
%(fairly significant) revisions. In particular, the proof of Theorem 3 is
%difficult to read due to the notation, and the paper could be substantially
%streamlined without losing the above results. Detailed suggestions follow.
%
%Major comments:

@@

\begin{response}
  Since the initial submission, there is now even more justification for including this special case.  A result of Bose et al on vertex ranking relies critically on the simple treewidth of $H$.  In particular, if $H$ is planar and has treewidth $3$, then $H\boxtimes P\boxtimes K_\ell$ an $r$-vertex ranking using $O(\log n/\log\log\log n)$ colours (for any fixed $r$ and $\ell$).  The three logs in the denominator come from the fact that $H$ has simple treewidth $3$.  Simple treewidth $4$ or even (non-simple) treewidth $3$ lead to more logs in the denominator.  The number of logs in the denominator is exactly the simple treewidth of $H$, and this is tight.

  Also, while adding details requested by Referee \#2 we discovered that a small change in the proof (removing both edges of a crossing pair instead of just one) results in a better constant and leads to a more general result that applies to $d$-framed graphs. This, in turn, leads to better results for $d$-map graphs.
\end{response}

4. Sections 4 and 5: I think as they are currently organized they distract from the purpose of the paper, as stated in the introduction, to ``prove product structure theorems for several non-minor-closed classes of interest.'' Which of Corollaries 1--4 and Theorems 12--17 are most important? I would remove any mention of the applications from the examples section, put that section first, and state in that section a single main theorem with all of the best bounds on the product structure. The applications section is mostly a survey and I think should be focused on the most exciting new corollaries. Theorems 18 and 19 should be in the applications section.

\textcolor{blue}{We have reorganised the paper, addressing this comment. }

@@

12. page 11, first paragraph of Section 3: ``This section applies our main results for shortcut systems to prove graph product structure theorem...'' $\longrightarrow$ ``to prove a graph product structure theorem...''

\done

\subsection*{Referee \#2}

%The Graph Product Structure Theorem of Dujmović, Joret,
%Micek, Morin, Ueckerdt, and Wood (FOCS '19) asserts that every planar
%graph is a subgraph of the strong product of a graph of treewidth at
%most 8 and a path.  This has been the key tool in resolving several
%longstanding open problems on planar graphs (and graphs in
%minor-closed classes), including problems on queue layouts,
%non-repetitive colouring, $p$-centered colouring, and universal graphs.
%This paper extends the Graph Product Structure Theorem beyond
%minor-closed classes.  The authors prove product structure theorems
%for k-planar graphs, map graphs, string graphs, graph powers, and
%nearest neighbour graphs.   The authors also present several
%applications of their results.  These are essentially the same
%applications as for the original Graph Product Structure Theorem; the
%point being that the applications typically follow quite easily once a
%product structure theorem is established.
%
%This paper is a significant step in answering the (ambitious) question
%of which graph classes admit product structure theorems. It is
%extremely well-written (with very few typos) and was a pleasure to
%read.  It is certainly well within the scope of JCTB and I strongly
%recommend it for acceptance.
%
%Below are some typos and suggestions for the authors.

@@

Page 13.  Regarding the proof of Lemma 4, I think it is better to
include all the details from reference [14].  As far as I know, JCTB
does not have a page limit, so I do not see an issue with just
reproducing the entire proof and telling the reader that they can skip
all the details if they wish.  This is just my personal opinion
though, so the authors can ignore this request if they choose.

\textcolor{blue}{Upon reflection we agree with the referee here. Indeed, while cleaning up this proof we have also improved the constant from $30$ to $7$. The proof is quite similar to the previous proof except that graph $G'$ is obtained by removing both crossing edges involved in a crossing (the spars of a kite), rather than just one.  We now use the language of framed multigraph, for consistency with the literature. Thanks to the referee for forcing us to write this proof more carefully so that we could discover this improvement.}

Page 15. Regarding the proof of Theorem 11, if it really is the same
proof as the proof of Theorem 2, why not just prove Theorem 11 (from
which Theorem 2 follows as a special case)?

\textcolor{blue}{We have written the paper starting with simpler cases before moving onto the more general. This approach may use a few more words, but we think the paper becomes more accessible for more readers this way.}

### 2022-03-24 k-planar.tex
minor edits and comments

For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it has an embedding in $\R^2$ having a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$. The graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.

@@

%\note{PM}{Consider a change of notation:   Replace $F$ with $C$.  (I was using $F$ before to be consistent with the previous section, but that notation is gone now.)} \note{DW}{I agree.}

@@

% \note{DW}{We need to define ``closure''}

\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ be a $d$-framed multigraph and let $G'$ be a frame of $G$.
		\item Let $F_0$ be the outer face of $G'$ and let $T$ be a BFS spanning forest of $G'$ rooted at $V(F_0)$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(v,V(F_0))=j\}$.
		\item Let $F$ be a cycle in $G'$ that $V(F)$ is partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is connected; and
			\item $P_i$ has a partition $\{X_i,Y_i\}$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the interior or boundary of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
		\item For each $x\in V(H)$, $S_x$ has a partition $\{X_x,Y_x\}$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
		\item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
	\end{compactenum}
%\note{PM}{I've changed my mind about this notation change.  We end up applying the lemma inductively on the faces $F_1,\ldots,F_m$ of $M$ (which are cycles in $G$).  It makes sense that the original cycle in $G$ is called $F$.}
%\note{DW}{okay}
%\begin{lem}
%  \label{induction} The setup:
%  \begin{compactenum}
%    \item Let $G$ be a $d$-framed multigraph and let $G'$ be a frame of $G$.
%    \item Let $F_0$ be the outer face of $G'$ and let $T$ be a BFS spanning forest of $G'$ rooted at $V(F_0)$.
%    \item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(v,V(F_0))=j\}$.
%    \item Let $F$ be a cycle in $G'$ whose vertices are partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
%    \begin{compactenum}
%      \item $F[P_i]$ is connected; and
%      \item $P_i=Y_i\cup X_i$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
%    \end{compactenum}
%    \item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
%  \end{compactenum}
%  Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
%  \begin{compactenum}[(i)]
%    \item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
%    \item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
%    \item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
%  \end{compactenum}
\end{lem}

@@

\begin{proof}
Let $G$ be a $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}.
  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.
We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges will appear on the boundary of a common face.

\begin{figure}
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
        \includegraphics{figs/one_planar_example-3} &
        \includegraphics{figs/one_planar_example-4} \\
        $G$ & $G'$
     \end{tabular}
    \end{center}
    \caption{The graph $G'$ obtained by removing pairs of crossing edges from $G$ is a plane multigraph whose faces all have three or four sides.}
    \label{one_planar_example}
\end{figure}

@@

Consider some pair of edges $vw$ and $xy$ that cross at $p$.  The point $p$ is a vertex of $G_0$ that has degree $4$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$ and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three or four edges; see \cref{one_planar_example}(b).  Therefore $G$ is a $4$-framed graph with frame $G'$.
% Since, for any pair of crossing edges $\{vw,xy\}$ that we remove this way, $vxwy$ is a cycle in $G'$, $\dist_{G'}(v,w)\le 2$ and $\dist_{G'}(x,y)\le 2$.
% It follows immediately that, for any pair of vertices $a,b\in V(G)$, $\dist_{G'}(a,b)\le 2\dist_G(a,b)$.
\end{proof}

@@

\note{PM}{Got rid of genus-$g$ in the following. We use shortcuts for $(g,d)$-map graphs and now we don't have to define $(g,d)$-frame.}

\begin{lem}
\label{MewMapGraphLemma}
For all integers $g\geq 0$ and $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.
\end{lem}

>>>>

\note{DW}{I suggest we use the language of ``embedded graph'' here}
For any integer $d\ge 3$, a (multi)graph $G$ is \defin{$d$-framed} if it has an embedding in $\R^2$ having a biconnected plane spanning (multi)subgraph $G_0$, whose facial cycles each have length at most $d$,  and such that the interior of each edge in $E(G)\setminus E(G_0)$ is contained in the interior of some face of $G_0$. The graph $G_0$ is called the \defin{frame} of $G$. This definition (for simple graphs) was introduced by \citet{BDGGMR}.

@@

@@

\note{DW}{Some suggestions: Replace $G'$ by $F$ (the frame), replace $F_0$ by $C$ (the outer-cycle), replace $F_1,\dots,F_m$ by $C_1,\dots,C_m$. Use $F_1,\dots,F_m$ for the subgraph of $F$ inside $C_i$. Use $G_i$ for the subgraph of $G$ inside $C_i$.}

\begin{lem}
	\label{induction} The setup:
	\begin{compactenum}
		\item Let $G$ be a $d$-framed multigraph and let $G'$ be a frame of $G$.
		\item Let $F_0$ be the outer face of $G'$ and let $T$ be a BFS spanning forest of $G'$ rooted at $V(F_0)$.
		\item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(v,V(F_0))=j\}$.
		\item Let $F$ be a cycle in $G'$ that $V(F)$ is partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
		\begin{compactenum}
			\item $F[P_i]$ is connected; and
			\item $P_i$ has a partition $\{X_i,Y_i\}$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
		\end{compactenum}
		\item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the interior or boundary of $F$.
	\end{compactenum}
	Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
	\begin{compactenum}[(i)]
		\item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
		\item For each $x\in V(H)$, $S_x$ has a partition $\{X_x,Y_x\}$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
		\item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$ then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
	\end{compactenum}
%\note{PM}{I've changed my mind about this notation change.  We end up applying the lemma inductively on the faces $F_1,\ldots,F_m$ of $M$ (which are cycles in $G$).  It makes sense that the original cycle in $G$ is called $F$.}
%\note{DW}{okay}
%\begin{lem}
%  \label{induction} The setup:
%  \begin{compactenum}
%    \item Let $G$ be a $d$-framed multigraph and let $G'$ be a frame of $G$.
%    \item Let $F_0$ be the outer face of $G'$ and let $T$ be a BFS spanning forest of $G'$ rooted at $V(F_0)$.
%    \item For every integer $j\ge 0$, let $L_j=\{v\in V(G):\dist_T(v,V(F_0))=j\}$.
%    \item Let $F$ be a cycle in $G'$ whose vertices are partitioned into $k\le 3$ sets $P_1,\ldots,P_k$ such that for each $i\in\{1,\ldots,k\}$,
%    \begin{compactenum}
%      \item $F[P_i]$ is connected; and
%      \item $P_i=Y_i\cup X_i$ where $|X_i|\le d-3$ and $|Y_i\cap L_j| \le 3$ for each $j\ge 0$.
%    \end{compactenum}
%    \item Let $N$ and $N'$ be the subgraphs of $G$ and $G'$ consisting only of those edges and vertices contained in the closure of $F$.
%  \end{compactenum}
%  Then $N$ has an $H$-partition $\PP=\{S_x: x\in V(H)\}$ such that:
%  \begin{compactenum}[(i)]
%    \item for each $i\in\{1,\ldots,k\}$, there exists $x_i\in V(H)$ such that $P_i=S_{x_i}$;
%    \item For each $x\in V(H)$, $S_x=Y_x\cup X_x$ where $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each $j\ge 0$;
%    \item $H$ has a $3$-simple tree decomposition $\mathcal{T}$ and, if $k=3$, then $\{x_1,x_2,x_3\}$ is contained in exactly one bag of $\mathcal{T}$.
%  \end{compactenum}
\end{lem}

@@

\begin{proof}
Let $G$ be a $1$-plane multigraph. We may assume that no two edges incident to a common vertex of $G$ cross, since such a crossing can be removed by a local modification to obtain an isomorphic 1-plane graph in which the two edges do not cross\footnote{While this is true for 1-plane graphs it is not true for $k$-plane graphs with $k\ge 3$; the uncrossing operation can increase the number of crossings on a particular edge from $k$ to $2(k-1)$.}.
  % We say that a vertex $v\in V(G)$ appears on a face $F$ if $v$ is contained in the closure of $F$.
We make $G$ into an \defin{edge-maximal} $1$-plane multigraph, by repeating the following operation:  If two vertices $v$ and $w$ of $G$ appear on a common face $F$ and there is no edge $vw\in E(G)$ contained in the boundary of $F$, then add the edge $vw$, embedded in the face $F$.  This may introduce parallel edges into $G$, but no two such edges  appear on the boundary of a common face.

\begin{figure}[h]
  \begin{center}
    \begin{tabular}{c@{\hspace{1cm}}c}
        \includegraphics{figs/one_planar_example-3} &
        \includegraphics{figs/one_planar_example-4} \\
        $G$ & $G'$
     \end{tabular}
    \end{center}
    \caption{The graph $G'$ obtained by removing pairs of crossing edges from $G$ is a plane multigraph whose faces all have three or four sides.}
    \label{one_planar_example}
\end{figure}

@@

Consider some pair of edges $vw$ and $xy$ that cross at point $p$, which a degree-4 vertex of $G_0$.  There are four faces $F_1,\ldots,F_4$ of $G_0$ that contain $p$ and, from the preceding discussion each of $vx$, $xw$, $wy$, and $yv$ is an uncrossed edge of $G$.  Therefore, by removing each pair of crossing edges from $G$ we obtain a plane multigraph $G'$ each of whose faces is bounded by three or four edges; see \cref{one_planar_example}(b).  Therefore $G$ is a $4$-framed graph with frame $G'$.
% Since, for any pair of crossing edges $\{vw,xy\}$ that we remove this way, $vxwy$ is a cycle in $G'$, $\dist_{G'}(v,w)\le 2$ and $\dist_{G'}(x,y)\le 2$.
% It follows immediately that, for any pair of vertices $a,b\in V(G)$, $\dist_{G'}(a,b)\le 2\dist_G(a,b)$.
\end{proof}

@@

%\note{PM}{Got rid of genus-$g$ in the following. We use shortcuts for $(g,d)$-map graphs and now we don't have to define $(g,d)$-frame.}

\begin{lem}
\label{MewMapGraphLemma}
For every %all 
integer %s $g\geq 0$ and
 $d\geq 3$, every $d$-map graph is a spanning subgraph of a $d$-framed graph.
\end{lem}
